He-Man: A domain-specific language for lightweight concurrency

### 2011-12-13 report/report.tex
Wirte a lot for the frontend.

\documentclass[preprint,11pt]{sigplanconf}
\usepackage[colorlinks=true,linkcolor=black,urlcolor=black,citecolor=black]%
{hyperref}
\usepackage{graphicx}
\renewcommand{\t}{\texttt}
\renewcommand{\b}{\textbf}
\renewcommand{\i}{\textit}

@@

\begin{figure}[ht]
\centering
\bnf{
\i{Stmt}
\p{\b{Decl} (\i{Var}, \i{Type}) \i{Expr}}
\a{\b{While} \i{Expr} [\i{Stmt}]}
\a{\b{If} \i{Expr} [\i{Stmt}] [\i{Stmt}]}
\a{\b{Spawn} ([(\i{Var}, \i{Type})], [\i{Stmt}]) [\i{Expr}]}
\a{\b{Assign} \i{Expr} \i{Expr}}
\a{\b{Exp} \i{Expr}}
\a{\b{Wait} \i{Expr}}
\a{\b{Exit}}
\i{Expr}
\p{\b{Call} \i{String} [\i{Expr}]}
\a{\b{Arith} \i{ArithOp} \i{Expr} \i{Expr}}
\a{\b{ArithUnop} \i{ArithUnop} \i{Expr}}
\a{\b{RelnOp} \i{RelnOp} \i{Expr} \i{Expr}}
\a{\b{Constant} \i{String}}
\a{\b{NumLit} \i{Integer}}
\a{\b{StringLit} \i{String}}
\a{\b{Var} \i{String}}
\a{\b{CurThread}}
\i{Type} 
\p{\b{Int} $\mid$ \b{Bool} $\mid$ \b{String}
   $\mid$ \b{Buffer} $\mid$ \b{Event}}
}
\caption{Front-end grammar.}
\end{figure}

>>>>

\documentclass[preprint,11pt]{sigplanconf}
\usepackage[colorlinks=true,linkcolor=black,urlcolor=black,citecolor=black]%
{hyperref}
\usepackage{graphicx}
\usepackage{verbatim}
\newenvironment{smallverbatim}{\endgraf\small\verbatim}{\endverbatim} 
\renewcommand{\t}{\texttt}
\renewcommand{\b}{\textbf}
\renewcommand{\i}{\textit}

@@

The underlying abstract syntax for the front-end He-Man language is a
fairly conventional imperative, statement based, first order language,
shown in Figure \ref{fig:front-syntax}. The notable constructs are
\b{Spawn}, which creates a new thread that takes a list of arguments
and launches the associated code with those arguments available to it,
and \b{Wait}, which takes an event expression as an argument and
blocks until the event has ``completed''. Note that \b{Call} is for
calling foreign functions; He-Man does not have its own notion of
function.

Since writing programs using the abstract syntax would be highly
inconvenient, we provide a monadic library for writing He-Man
programs, which allows us to leverage Haskell's do notation to write
He-Man programs using convenient imperative syntax. In this way, we
are taking advantage of monads being ``programmable semi-colons''.

We define the \t{Prog} monad, which contains a \t{State} monad used to
maintain a counter for generating unique variable names and a
\t{Writer} monad that is used to build up a list of statements. We
then provide \t{Prog} wrappers for statements that add the statement
to the list of statements being built up by the \t{Writer}. The
wrappers for \b{Spawn}, \b{Assign}, \b{Wait}, and \b{Exit} simply take
the obvious arguments, construct a \t{Stmt} and add it to the writer
list. The \b{Assign} wrapper is an infix operator, and is named
``.=''. The wrappers for \b{While} and \b{If} take values in \t{Prog}
instead of statement lists, and run the \t{Writer} to extract the
lists of statements. The monadic wrapper for \b{Decl} declares a
variable (with a name made unique by appending a unique number) and
then returns the \t{Expr} that references the declared variables. This
makes declaring and using variables very natural. We also provide a
wrapper for \b{Call} that declares a fresh variable, assigns the
result of a call to it, and then returns the variable. Then,
``declaring'' a foreign function consists of writing a convenient
wrapper function that invokes the \b{Call} wrapper.

At the expression level, \t{Expr} is an instance of the \t{Num} type
class so that numeric literals and operations can be used in He-Man
syntax. We also provide infix operators for other useful operations on
expressions, some of them with mildly awkward names, since the good
names were already taken by a function with the wrong type.

The result of all of this is that He-Man code can be written in a
fairly natural style, as shown in Figure \ref{fig:code-ex}.

%We also provide an \b{Assign} wrapper that takes a \t{Prog Expr} as its argument

\begin{figure}[ht]
\centering
\bnf{
\i{Stmt}
\p{\b{Decl} (\i{Var}, \i{Type}) \i{Expr}}
\a{\b{While} \i{Expr} [\i{Stmt}]}
\a{\b{If} \i{Expr} [\i{Stmt}] [\i{Stmt}]}
\a{\b{Spawn} ([(\i{Var}, \i{Type})], [\i{Stmt}]) [\i{Expr}]}
\a{\b{Assign} \i{Expr} \i{Expr}}
\a{\b{Exp} \i{Expr}}
\a{\b{Wait} \i{Expr}}
\a{\b{Exit}}
\i{Expr}
\p{\b{Call} \i{String} [\i{Expr}]}
\a{\b{Arith} \i{ArithOp} \i{Expr} \i{Expr}}
\a{\b{ArithUnop} \i{ArithUnop} \i{Expr}}
\a{\b{RelnOp} \i{RelnOp} \i{Expr} \i{Expr}}
\a{\b{Constant} \i{String}}
\a{\b{NumLit} \i{Integer}}
\a{\b{StringLit} \i{String}}
\a{\b{Var} \i{String}}
\a{\b{CurThread}}
\i{Type} 
\p{\b{Int} $\mid$ \b{Bool} $\mid$ \b{String}
   $\mid$ \b{Buffer} $\mid$ \b{Event}}
}
\caption{Front-end grammar.}
\label{fig:front-syntax}
\end{figure}

\begin{figure}[ht]
\centering
\begin{smallverbatim}
child_code = declare_thread [("child_fd", Int)] $
  \[child_fd] -> do
  ev <- setup_connection child_fd
  buf <- new_buf bufsize
  while 1 $ do
    amt_read <- do_read ev buf bufsize
    ifthen (amt_read .== 0) exit
    amt_written <- full_write ev buf amt_read
    ifthen (amt_written .< amt_read) exit
\end{smallverbatim}
\caption{He-Man example from echo server.}
\label{fig:code-ex}
\end{figure}
